A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2015-10-06 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Fix broken line breaks

The increase in formality of the ontological argument has required a development of its basic notions. G\"odel's notion of positive property and Leibniz's notion of perfection differ. A formal distinction is that Leibniz's perfections are atomic whereas G\"odel's positive properties can consist of combinations of atomic properties \citep{fitting}[p.139]. In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being God-like is positive. Intuitively, a (possibly infinite) conjunction of positive properties is deduced from the universal definition of God-likeness. This deductive inference is not formalizable in a finite first-order calculus. The interplay between universal quantification (in the definition of a God-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) could explain why, starting with Scott \cite{scott}, this axiom o!
 f G\"odel has been replaced by another that simply assumes the positivity of the property of being god-like.

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \cite{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \cite{hajek}). Nevertheless, some authors of variants \cite[footnotes !
 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of assuming constant domains is motivated by simplicity.

@@

Scott's version of G\"odel's proof depends on 5 axioms that circumscribe the notion of \emph{positive} property, with positivity being denoted by the undefined second-order predicate symbol $P$. Additionally, 3 definitions are used for defining the notions of \emph{God-like}, \emph{essence} and \emph{necessary existence}. Technically, these definitions only abbreviate certain complex formulas. The argument would still go through if all defined symbols were replaced by the complex formulas they define. This observation is particularly relevant in the case of \emph{necessary existence}. Since this notion of ``existence'' is just an abbreviation, G\"odel's argument is not susceptible to Kant's criticism against Anselm's argument (that existence should not be treated as a predicate). In G\"odel's proof, existence is properly denoted by the existential quantifier. It is, therefore, unfortunately misleading to refer to the defined predicate symbol $E$ as ``necessary existence'', w!
 hen in fact it is just a convenient abbreviation.

@@

Although the simpler proof of Theorem 1 presented here was developed independently, it seems conceptually related to a possible simplification briefly and informally described by Anderson \cite[footnote 2]{and90} and attributed by him to an anonymous referee. The simpler proof of Theorem 1 is not particularly special. In fact, it is probably the most likely proof that any well-trained logician would construct, if asked to derive Theorem 1 from Axioms 1 and 2 using methodical formal proof search techniques. Modern techniques usually rely on unification to compute or guess good instances for weakly quantified variables, and therefore they would never result in instantiations containing $=$ and $\neq$, which are symbols that do not appear anywhere in the axioms or in the theorems to be proved. We may only wonder why G\"odel preferred to use those unusual instantiations. It might be an interesting question from a historical point of view, if this suggests that there might have b!
 een a change (perhaps evolution) in the way logicians construct formal proofs; or even from a philosophical point of view, if G\"odel's preference was philosophically motivated.

>>>>

The increase in formality of the ontological argument has required a development of its basic notions. G\"odel's notion of positive property and Leibniz's notion of perfection differ. A formal distinction is that Leibniz's perfections are atomic whereas G\"odel's positive properties can consist of combinations of atomic properties \citep{fitting}[p.139]. In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being God-like is positive. Intuitively, a (possibly infinite) conjunction of positive properties is deduced from the universal definition of God-likeness. This deductive inference is not formalizable in a finite first-order calculus. The interplay between universal quantification (in the definition of a God-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) could explain why, starting with Scott \cite{scott}, this axiom of G\"odel has been replaced by another that simply assumes the positivity of the property of being god-like.

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \cite{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \cite{hajek}). Nevertheless, some authors of variants \cite[footnotes 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of assuming constant domains is motivated by simplicity.

@@

Scott's version of G\"odel's proof depends on 5 axioms that circumscribe the notion of \emph{positive} property, with positivity being denoted by the undefined second-order predicate symbol $P$. Additionally, 3 definitions are used for defining the notions of \emph{God-like}, \emph{essence} and \emph{necessary existence}. Technically, these definitions only abbreviate certain complex formulas. The argument would still go through if all defined symbols were replaced by the complex formulas they define. This observation is particularly relevant in the case of \emph{necessary existence}. Since this notion of ``existence'' is just an abbreviation, G\"odel's argument is not susceptible to Kant's criticism against Anselm's argument (that existence should not be treated as a predicate). In G\"odel's proof, existence is properly denoted by the existential quantifier. It is, therefore, unfortunately misleading to refer to the defined predicate symbol $E$ as ``necessary existence'', when in fact it is just a convenient abbreviation.

@@

Although the simpler proof of Theorem 1 presented here was developed independently, it seems conceptually related to a possible simplification briefly and informally described by Anderson \cite[footnote 2]{and90} and attributed by him to an anonymous referee. The simpler proof of Theorem 1 is not particularly special. In fact, it is probably the most likely proof that any well-trained logician would construct, if asked to derive Theorem 1 from Axioms 1 and 2 using methodical formal proof search techniques. Modern techniques usually rely on unification to compute or guess good instances for weakly quantified variables, and therefore they would never result in instantiations containing $=$ and $\neq$, which are symbols that do not appear anywhere in the axioms or in the theorems to be proved. We may only wonder why G\"odel preferred to use those unusual instantiations. It might be an interesting question from a historical point of view, if this suggests that there might have been a change (perhaps evolution) in the way logicians construct formal proofs; or even from a philosophical point of view, if G\"odel's preference was philosophically motivated.
